---
title: "Class 11"
author: "Achyuta Ramesh - PID A16956100"
format: gfm
---

```{r}
file <- read.csv("https://bioboot.github.io/bggn213_W19/class-material/rs8067378_ENSG00000172057.6.txt")

```

```{r}
table <- read.table("https://bioboot.github.io/bggn213_W19/class-material/rs8067378_ENSG00000172057.6.txt")

table(table$geno)
```

```{r}
library(ggplot2)
boxplot <- ggplot(table) + aes(geno, exp, fill = geno) + 
  geom_boxplot(notch = TRUE) + 
  stat_summary()
summary(boxplot)
boxplot
```

```{r}
boxplot
```



```{r}
library(dplyr)

AGs <- table |>
  filter(geno == "A/G") 

median(AGs$exp)

AAs <- table |>
  filter(geno == "A/A") 

median(AAs$exp)

GGs <- table |>
  filter(geno == "G/G") 

median(GGs$exp)
```
> Q13: Read this file into R and determine the sample size for each genotype and their
corresponding median expression levels for each of these genotypes.
There are 108 A/As, 233 A/Gs, and 121 G/Gs. The median expression levels for these, respectively, are 31.24, 25.06, and 20.07.
Q14: Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?
The expression of A/A is significantly higher than that of G/G, so I can conclude that the A/A genotype leads to higher expression of ORMDL3 than the G/G one does. Therefore, the SNP does affect expression of ORMDL3.
