---
title: "Class 15 - Pertusssis Mini Project"
author: "Achyuta"
format: html
---

Pertussis data by year:

[CDC Data](https://www.cdc.gov/pertussis/php/surveillance/pertussis-cases-by-year.html)

We will use the datapasta R package to "scrape" this data into R.

```{r}
cdc <- data.frame(
                                 Year = c(1922L,1923L,1924L,1925L,
                                          1926L,1927L,1928L,1929L,1930L,1931L,
                                          1932L,1933L,1934L,1935L,1936L,
                                          1937L,1938L,1939L,1940L,1941L,1942L,
                                          1943L,1944L,1945L,1946L,1947L,
                                          1948L,1949L,1950L,1951L,1952L,
                                          1953L,1954L,1955L,1956L,1957L,1958L,
                                          1959L,1960L,1961L,1962L,1963L,
                                          1964L,1965L,1966L,1967L,1968L,1969L,
                                          1970L,1971L,1972L,1973L,1974L,
                                          1975L,1976L,1977L,1978L,1979L,1980L,
                                          1981L,1982L,1983L,1984L,1985L,
                                          1986L,1987L,1988L,1989L,1990L,
                                          1991L,1992L,1993L,1994L,1995L,1996L,
                                          1997L,1998L,1999L,2000L,2001L,
                                          2002L,2003L,2004L,2005L,2006L,2007L,
                                          2008L,2009L,2010L,2011L,2012L,
                                          2013L,2014L,2015L,2016L,2017L,2018L,
                                          2019L,2020L,2021L,2022L),
         No..Reported.Pertussis.Cases = c(107473,164191,165418,152003,
                                          202210,181411,161799,197371,
                                          166914,172559,215343,179135,265269,
                                          180518,147237,214652,227319,103188,
                                          183866,222202,191383,191890,109873,
                                          133792,109860,156517,74715,69479,
                                          120718,68687,45030,37129,60886,
                                          62786,31732,28295,32148,40005,
                                          14809,11468,17749,17135,13005,6799,
                                          7717,9718,4810,3285,4249,3036,
                                          3287,1759,2402,1738,1010,2177,2063,
                                          1623,1730,1248,1895,2463,2276,
                                          3589,4195,2823,3450,4157,4570,
                                          2719,4083,6586,4617,5137,7796,6564,
                                          7405,7298,7867,7580,9771,11647,
                                          25827,25616,15632,10454,13278,
                                          16858,27550,18719,48277,28639,32971,
                                          20762,17972,18975,15609,18617,
                                          6124,2116,3044)
       )
```

```{r}
library(ggplot2)
baseplot <- ggplot(cdc) +
  aes(x = Year, y = No..Reported.Pertussis.Cases) + 
  geom_point() + 
  geom_line()
```


Add some landmark developments as annotation to our plot. We include the first whole-cell (wP) vaccine roll-out in 1940. 

Let's add the switch to acellular vaccine (aP) in 1996. 

```{r}
baseplot + 
  geom_vline(xintercept = 1946, col = "blue") + 
  geom_vline(xintercept = 1995, col = "pink") + 
  geom_vline(xintercept = 2020, col = "purple") + 
  geom_vline(xintercept = 2004, col = "light green")
```

We went from ~200,000 cases pre wP vaccine to ~1,000 cases in 1976. The US switched to the aP vaccine in 1995. We start to see a big increase in 2004 to ~26,000 cases.

There is a ~10 year lag from aP roll-out to increasing case numbers. This holds true of other countries like Japan, UK, etc.

**Key queestion**: Why does the aP vaccine induced immunity wane faster than that of the wP vaccine?

##CMI-PB

The CMI-PB (computational models of Immunity Pertussis Boost) makes available lots of data about the immune response to Pertussis booster vaccination.

Critically, it tracks wP and aP individuals over time to see how their immune response changes. 

CMI-PB makes all their data freely available via JSON format tables from their database.

Let's read the first one of these tables.

```{r}
library(jsonlite)

subject <- read_json("https://www.cmi-pb.org/api/v5/subject",
                     simplifyVector = TRUE)

head(subject)
```

> Q1. How many subjects are there in this datase?

```{r}
nrow(subject)
```

> Q2. How many aP and wP individuals are there?

```{r}
table(subject$infancy_vac)
```

> Q3. How many males and females are there?

```{r}
table(subject$biological_sex)
```

> Q4. Breakdown by biological sex and race.

```{r}
table(subject$race, subject$biological_sex)
```

> Q Does this do a good job of representing the US populus?

No

Let's get more data from CMI-PB, this time about the specimens collected.

```{r}
specimen <- read_json("https://www.cmi-pb.org/api/v5/specimen",
                      simplifyVector = TRUE)

head(specimen)
```

Now we can merge these two tabless `subject` and `specimen` to make one new `meta` table with the combined data.

```{r}
library(dplyr)

meta <- inner_join(subject, specimen)

head(meta)
```

Now read an "experiment data" table from CMI-PB

```{r}
abdata <- read_json("https://www.cmi-pb.org/api/v5/plasma_ab_titer",
                    simplifyVector = TRUE)

head(abdata)
```

One more join to do of `meta` and `abdata` to  associate all the metadata about the individual and their race, biological sex, and infancy vaccination status together with antibody levels

```{r}
ab <- inner_join(meta, abdata)
head(ab)
```

```{r}
nrow(ab)
```

How many isotypes

```{r}
table(ab$isotype)
```

How many antigens?
```{r}
table(ab$antigen)
```

Let's focus in on IgG - one of the main antibody types responsive to bacterial or viral infections.

```{r}
igg <- filter(ab, isotype == "IgG")
head(igg)
```

Make a first plot of MFI (Mean Fluorescence Intensity - measure of how much is detected) for each antigen.

```{r}
ggplot(igg) + 
  aes(MFI_normalised, antigen) + 
  geom_boxplot()
```

```{r}
ggplot(igg) + 
  aes(MFI_normalised, antigen, col = infancy_vac) + 
  geom_boxplot() + 
  facet_wrap(~visit)
```


```{r}
table(igg$visit)
```

Looks like we don't have data yet for all subjects in terms of visits 8 onwards. So let's exclude these. 

```{r}
igg_7 <- filter(igg, visit %in% 1:7)
table(igg_7$visit)
```

```{r}
ggplot(igg_7) + 
  aes(MFI_normalised, antigen) + 
  geom_boxplot() + 
  facet_wrap(~visit)
```

Let's try a different plot. First focus on one antigen, start with PT (Pertussis Toxin) and plot visit or time on the x-axis and MFI normalized on the y axis.

```{r}
ggplot(igg_7) + 
  aes(visit, MFI_normalised) +
  geom_point()
```

```{r}
abdata.21 <- ab %>% filter(dataset == "2021_dataset")

abdata.21 %>% 
  filter(isotype == "IgG",  antigen == "PT") %>%
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```

```{r}
url <- "https://www.cmi-pb.org/api/v2/rnaseq?versioned_ensembl_gene_id=eq.ENSG00000211896.7"

rna <- read_json(url, simplifyVector = TRUE) 
```

```{r}
meta <- inner_join(specimen, subject)
ssrna <- inner_join(rna, meta)
```

```{r}
ggplot(ssrna) +
  aes(x = visit, y = tpm, group=subject_id) +
  geom_point() +
  geom_line(alpha=0.2)
```

```{r}
ssrna %>%  
  filter(visit==4) %>% 
  ggplot() +
    aes(tpm, col=infancy_vac) + geom_density() + 
    geom_rug() 
```

